Human capital mediates natural selection in contemporary humans

### 2022-03-14 response-to-reviewers-BE.Rmd

---
title: "Response to reviewers"
author: "David Hugh-Jones and Abdel Abdellaoui"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: 
  pdf_document:
    toc: false
    latex_engine: xelatex
header-includes:
  - \usepackage{subfig}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
---

```{r echo=FALSE}

suppressWarnings(suppressPackageStartupMessages(library(dplyr)))
library(ggplot2)
library(forcats)

drake::loadd(res_edu)

theme_set(theme_minimal())

order_abs <- function (n = 1) {
  function (x) x[n] 
}

pretty_names <- function (names) {
  pretty <- c(
    ADHD_2017               = "ADHD",
    age_at_menarche         = "Age at menarche",
    age_at_menopauze        = "Age at menopause",
    agreeableness           = "Agreeableness",
    ai_substance_use        = "Age at smoking initiation",
    alcohol_schumann        = "Alcohol use",
    alzheimer               = "Alzheimer",
    autism_2017             = "Autism",
    bipolar                 = "Bipolar",
    bmi_combined            = "BMI",
    body_fat                = "Body Fat",
    caffeine                = "Caffeine",
    cannabis                = "Cannabis (ever vs. never)",
    cognitive_ability       = "Cognitive Ability",
    conscientiousness       = "Conscientiousness",
    coronary_artery_disease = "Coronary Artery Disease",
    cpd_substance_use       = "Cigarettes per day",
    diagram_T2D             = "Type 2 Diabetes",
    dpw_substance_use       = "Drinks per week",
    EA2_noUKB               = "Educ. attainment 2 (no UKBB)",
    EA3_excl_23andMe_UK     = "Educ. attainment 3 (no UK)",
    eating_disorder         = "Eating disorder",
    extraversion            = "Extraversion",
    height_combined         = "Height",
    hip_combined            = "Hip circumference",
    MDD_PGC2_noUKB          = "Major Depressive Disorder",
    neuroticism             = "Neuroticism",
    openness                = "Openness",
    sc_substance_use        = "Smoking cessation",
    SCZ2                    = "Schizophrenia",
    si_substance_use        = "Smoking initiation",
    wc_combined             = "Waist circumference",
    whr_combined            = "Waist-hip ratio"
  )

  pretty[names]
}

add_n <- function (var, ...) {
  var <- as.factor(var)
  n <- c(88274, 99987, 154603)
  levels(var) <- sprintf("%s (N = %s)", levels(var), 
                         prettyNum(n[seq_len(nlevels(var))], big.mark = ","))
  # var_n <- sprintf("%s (N = %s)", var, prettyNum(n[var], big.mark = ","))
  # var_n[is.na(var)] <- NA
  var
}

my_hline <- geom_hline(yintercept = 0, colour = "grey20", linetype = "dotted")
my_vline <- geom_vline(xintercept = 0, colour = "grey20", linetype = "dotted")
pval_palette <- c("FALSE" = "grey65", "TRUE" = "#FAC32D")

standard_ggplot <- function (
  dfr, 
  fill_col, 
  n_regs, 
  ...,
  score_col = quo(term), 
  order_idx = 1, 
  fill_direction = 1,
  n_cats    = NULL,
  conf_int  = NULL
) {
  if (! missing(score_col)) score_col <- enquo(score_col)

  mfc <- missing(fill_col)
  fill_col <- if (mfc) quo(NULL) else enquo(fill_col)
  if (missing(n_cats)) {
    n_cats <- if (mfc) 1 else length(unique(pull(dfr, {{fill_col}})))
  }

  if (is.null(conf_int)) conf_int <- "conf.low" %in% names(dfr)

  n_regs <- as.double(n_regs)

  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL

  dfr %>% mutate(
    !! score_col := pretty_names(!! score_col),
    !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
  ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
               fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
    my_vline + 
    confint_geom_segment +
    pval_geom_point(fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
    pval_scale() +
    my_fill_scale(n = n_cats, direction = fill_direction) +
    my_labs() +
    my_theme() +
    scale_linetype_manual(name = "", values = c("95% c.i." = 1))
}

shapes_ggplot <- function (
  dfr, 
  fill_col, 
  n_regs, 
  ...,
  score_col = quo(term), 
  order_idx = 1, 
  fill_direction = 1,
  n_cats    = NULL,
  conf_int  = NULL
) {
  if (! missing(score_col)) score_col <- enquo(score_col)

  mfc <- missing(fill_col)
  fill_col <- if (mfc) quo(NULL) else enquo(fill_col)
  if (missing(n_cats)) {
    n_cats <- if (mfc) 1 else length(unique(pull(dfr, {{fill_col}})))
  }

  if (is.null(conf_int)) conf_int <- "conf.low" %in% names(dfr)

  n_regs <- as.double(n_regs)

  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL

  dfr %>% mutate(
    !! score_col := pretty_names(!! score_col),
    !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
  ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
               shape = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
    my_vline + 
    confint_geom_segment +
    geom_point(size = 1.8, alpha = 0.8) +
    scale_color_manual(values = c("FALSE" = "grey45", "TRUE" = "orange"), ...) +
    scale_shape_manual(values = 1:n_cats + 14) +
    my_labs() +
    my_theme() +
    scale_linetype_manual(name = "", values = c("95% c.i." = 1))

}

tweaked_ggplot <- function (
  dfr, 
  fill_col, 
  n_regs, 
  ...,
  score_col = quo(term), 
  order_idx = 1, 
  fill_direction = 1,
  n_cats    = NULL,
  conf_int  = NULL
) {
  if (! missing(score_col)) score_col <- enquo(score_col)

  mfc <- missing(fill_col)
  fill_col <- if (mfc) quo(NULL) else enquo(fill_col)
  if (missing(n_cats)) {
    n_cats <- if (mfc) 1 else length(unique(pull(dfr, {{fill_col}})))
  }

  if (is.null(conf_int)) conf_int <- "conf.low" %in% names(dfr)

  n_regs <- as.double(n_regs)

  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL

  dfr %>% mutate(
    !! score_col := pretty_names(!! score_col),
    !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
  ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
               fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
    my_vline + 
    confint_geom_segment +
    pval_geom_point(size = 1.6, stroke = 0.8, fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
    scale_color_manual(values = c("FALSE" = "grey65", "TRUE" = "#f3b372")) +
    my_fill_scale(n = n_cats, direction = fill_direction) +
    my_labs() +
    my_theme() +
    scale_linetype_manual(name = "", values = c("95% c.i." = 1))
}

my_theme <- function (...) theme(axis.text.y = element_text(size = 7), ...)

pval_scale <- function (...) scale_color_manual(values = pval_palette, ...)

pval_geom_point <- function (size = 1.5, stroke = 0.5, fill = NULL, ...) {
  if (is.null(fill)) {
    geom_point(shape = 21, size = size, stroke = stroke, ...)
  } else {
    geom_point(shape = 21, size = size, stroke = stroke, fill = fill, ...)
  }
}

confint_geom_segment <- function () {
  geom_segment(aes(x = conf.low, xend = conf.high, linetype = "95% c.i."), 
               color = "grey60", alpha = 0.25, size = 1.1, 
               lineend = "round")
}

my_fill_scale <- function (n, aesthetics = "fill", ...) {

  fill_colors <- c("steelblue4", "#D63C3C")

  if (n <= 2) {
    dots <- list(...)
    dots$direction <- NULL
    do.call(scale_colour_manual, c(list(values = fill_colors[1:n],
                                        aesthetics = aesthetics), 
                                   dots))
  } else {
    scale_color_brewer(aesthetics = aesthetics, ...)
  }
}

my_labs <- function (x = "Effect size", y = "", ...) labs(x = x, y = y, ...)

n_regs <- as.double(nrow(res_edu))
res_edu <- res_edu %>% 
  mutate(
    "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
    "Age left FTE" = add_n(`Age left FTE`, "age_fte_cat")
  )

```

Dear Professor Keller,

Thank you very much for the chance to revise our manuscript. We would also like to thank both reviewers for their comments. Below we deal with the points they raised.

## Editor

> In particular, Reviewer 2 makes several good points that I believe can improve the clarity and flow of the manuscript and I suggest you consider these carefully. Reviewer 1 was highly favorable but also makes several excellent suggestions -- including pointing out that a similar observation has been made (remarkably) almost 100 years ago by Fisher. This seems important to include.

See below.

## Reviewer 1

> This is an outstanding paper, perhaps the most important that will be published in the field of behavioral genetics this year.

Thank you 😊.

> Most curious is the failure to cite Fisher (1930).... The overlap in theory and observations with the present work is striking, so much so that the omission is equally striking.

We thank the reviewer for giving us this reference. Indeed it is striking. Fisher's theory about the positive correlation he mentions is that it's driven by selection bias: Yale parents who can afford to send 6 children to Yale are "more able" than those who can only send 1, and as a result, among Yale students, children from bigger families are smarter. This is related to the income effect in our theory. We now mention this work in our conclusion. XXX maybe do more.

> The authors do not have to respond to my next point. This paper already has a lot. But I am curious as to whether the authors have an opinion about Fisher's proposed solution to this problem (if it is a problem). He in essence suggested that the population be stratified into classes based on "human capital," as we might say now, and that within each stratum redistribution should be performed so as to reward parents of larger families. The basic idea is that if I have four children, while an academic colleague who is the same age as me has none (which has allowed him to publish twice as many papers), the tax-and-spend system will take away some of his income and give it to me, the total amount of redistribution being such as to remove the fundamental tradeoff. Can this work in theory? Can it be made politically feasible?

Indeed, we don't want to talk about policy in the paper - we think we do not yet know enough. For what it's worth, here are my personal thoughts as corresponding author. First, government schemes to encourage fertility have a poor track record, see e.g. Gauthier (2007). Second, my strong gut feeling is that it is wrong for governments to (bluntly) institute breeding programs for their citizens. I wrote more informally on this topic [here](https://wyclif.substack.com/p/the-natural-selection-paper-part-822?s=w).

> Another general point is that there is some uncertainty over whether the fertility-IQ correlation is negative in entire populations. Some suggest that its negative value in unrepresentative samples is the result of ascertainment bias (e.g., Higgins, Reed, & Reed, 1962; Kolk & Kieron, 2019). In fact, I have reviewed a manuscript where the authors declined using the UK Biobank for a particular analysis (while using it for all others) on the grounds that the fertility-IQ relationship cannot be accurately inferred in that dataset. My position is that the fertility-IQ correlation has been negative, at most times since relevant data have been collected, while being less so at certain atypical times (e.g., the Baby Boom) and in men. In the present work, the authors use sample reweighting to address ascertainment bias in the UK Biobank, but this is not guaranteed to be effective. A brief literature review addressing the sign of the fertility-IQ correlation, in most Western populations in recent decades, therefore seems warranted to provide some reassurance. My suggestion is to begin with Lynn and Van Court (2004), Meisenberg (2010), and Reeve, Heeney, and Woodley (2018).

We agree this is an important topic, but we are more focused on human capital and earnings than IQ specifically. (Of course the polygenic score for educational attainment does also predict IQ.) The negative fertility-income relationship, in developed countries, is a very well-established stylized fact, and we mention some papers that also show income is causal.

Regarding ascertainment bias, the reviewer is right that there are no guarantees we can eliminate it. When we do correct for ascertainment bias, every correction increases our estimated effect sizes. So we expect that any remaining ascertainment also probably biases our estimates towards zero. We now point this out. (XXX say this somewhere?)

> abstract: "Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against)." This is hard to understand. How about: "Consistently over time, polygenic scores predicting higher earnings, education, and health also predict lower fertility."

Done.

> p\. 2: "natural selection may substantially increase the correlation between genetic scores and income, i.e. the "genetic lottery" (Harden, 2021)." Although the meaning eventually becomes clear, this sentence may be very cryptic.

We have made it more explicit: "natural selection may substantially increase the correlation between genetic scores and income, increasing genetic differences between different social groups, and thus making the "genetic lottery" (Harden 2021) more unfair."

> p\. 4, footnote 1: That the covariance between the number of children and the polygenic score for a given phenotype is equal to the evolutionary change in the phenotype between generations is, after some adjustment of terminology, Robertson's (1966, 1968) Secondary Theorem of Natural Selection (see also Walsh & Lynch, 2018, pp. 165-167). I think this should be pointed out.

We now cite Robertson (1966).

> p\. 4: It should be "stabilizing" selection, not "balancing" selection. The latter refers to any kind of selection that favors the coexistence of two or more alleles at a site. The former refers to selection that favors an intermediate value of a quantitative trait, and this is clearly what the authors mean. While I believe the term "diversifying" selection is sometimes uses for selection favoring both extremes of a quantitative trait, "disruptive" selection is probably more common. A similar point applies on p. 20.

We have changed vocabulary accordingly.

> p\. 10: I am interested in the functional form chosen for u(y). This has sensible properties for sigma \< 1 near 1, including being increasing in y for y somewhat greater than zero and being concave. Intuitively, this means that utility increases with income (net of costs of childrearing) but less so as income increases. Is there any particular reason for this functional form as opposed to others that also satisfy these properties? Also, is there any empirical evidence that "utility" obeys this relationship?

The functional form is a standard one in economics. It has the specific property that relative risk aversion is constant. That is, if you have CRRA utility and £100, and you are indifferent to a 50-50 bet between winning £30 and losing £20, then if you had £1000 you would be indifferent between a 50-50 to win £300 or lose £200.

The literature on estimating utility functions is vast (see e.g. Chuang and Schechter 2015). One issue is that of context. CRRA may make sense for risky financial decisions. Here, as Reviewer 2 points out, the "curvature" of the utility function is used not to explain attitudes to risk, but to trade off income against children at different income levels. This author wishes there was a single utility function that worked to explain all human behaviour across contexts... in reality, economists tend to use "what works". We use the CRRA form because it is tractable; we expect that using other forms, intermediate levels of risk aversion would still generate our empirical results, but the maths would be uglier. For a useful discussion of microeconomic theories of fertility, including functional forms, see Jones et al. (2008).

> p\. 30: I feel that more thought might be given to the choice of covariates. For example, if fluid IQ is on the causal path between the PGS and EA, then controlling it will bias downward the estimate of the the PGS effect on EA. But since this mediation analysis seems tentative and not crucial to the paper, perhaps some looseness here can be tolerated.

Yes. Conversely, fluid IQ might affect fertility choices via other pathways than human capital in the labour market, and not controlling for it might leave these to be picked up by EA. We prefer to be "conservative" in controlling for it, in the sense that we'd rather bias the estimate indirect effect of EA towards zero than away from zero. We now mention this issue in the appendix.

That said, the mediation analysis is really the central test of our theory. So we have tried to make it seem less tentative. See our response to Reviewer 2.

> p\. 34: For various reasons I think this section (and the bit in the main text that it supports) ought to be eliminated. The point is to show that natural selection increases the correlation between the EA PGS and income from one generation to the next, correcting for the noise in the PGS. I think a qualitative statement about this point is adequate; the more quantitative attempt by the authors has some problems.
>
> (1) "PSEA" is not defined, although I'm guessing this means "polygenic score for educational attainment."
>
> (2) It happens that there is a more recent meta-analysis of relevant twin studies (Silventoinen et al., 2020).
>
> (3) There appears to be substantial passive gene-environment correlation ("genetic nurture," some now call it), as a result of highly educated parents providing both favorable genes and a social status promoting educational attainment (e.g., Kong et al., 2018). I am sure this complicates the analysis in some way. The authors might respond that they deal with this by using within-family estimates, but see below.
>
> (4) The proof of Equation 5 uses the "standard errors-in-variables formula" plim \\hat\\beta = \\lambda\\beta. While this relationship seems plausible, it may not be standard to behavioral geneticists.
>
> (5) When there is genetic nurture, what is the right PGS effect to use in order to address the question here? The effect in the population, or the effect within families? If it is the latter, a problem is that the estimate of a 142.6% change in R\^2 between generations seems far too high and indicative of low statistical precision.
>
> (6) The assumption that eta is independent of p_star seems very strong. It includes, for example, the assumption that the PGS over rare variants is uncorrelated with the PGS over common variants. I think this is unlikely to be true, as a result of assortative mating (Yengo et al., 2018).

We accept point (6) in particular. On the other hand, we do state that this is a "back of the envelope calculation". We've had pushback from people saying "this is no big deal", so we want to show that at least under some assumptions, yes it might be. Note also that as we say:

"If unmeasured PSEA correlates positively with $\varepsilon$, then the reliability coefficient $\lambda$ will be lower than calculated. However, the proportional effect of natural selection on the PSEA-income relationship will remain the same, so long as $\lambda$ does not differ between the two sets of weighted regressions."

The within-family estimate should remove the worries about exogeneity of p\*, but they are estimated with a smaller sample. We agree that the estimate is high. The bootstrapped 95% confidence interval is from around 0% to 300%, which is not really informative, though it does suggest that genetic nurture is more likely to bias our results downwards than upwards. We've removed the discussion.

> p\. 40 and following: I confess to not following the derivations completely. The authors seem adept at what they do, and I trust that they have made no serious errors.

We hope so too.

### 2022-03-14 model-appendix.Rmd
Add proof that dN^*/da gets larger at higher levels of human capital
Response to BG reviewers: - add response document - move to RLRS not absolute N children   - some changes to over-time results - add analysis for respondents by Townsend deprivation - add bootstrap for effect of natural selection on income-EA3 R2, in FE model - add graph of mediation analysis - tweak plot style - other tweaks

Plugging (\ref{eq:Y1}) and (\ref{eq:Y2}) into the above, we can
derive the Karush-Kuhn-Tucker conditions for an optimum $(N_{1}^{*},N_{2}^{*},s^{*})$
as:
\begin{align}
\frac{d\mathcal{L}}{dN_{1}}=-bY_{1}^{-\sigma}+a+\lambda_{1} & =0\textrm{, with }\lambda_{1}=0\textrm{ if }N_{1}^{*}>0;\label{eq:kkt1}\\
\frac{d\mathcal{L}}{dN_{2}}=-bs^{*}hY_{2}^{-\sigma}+a+\lambda_{2} & -\lambda_{3}=0\textrm{, with }\lambda_{2}=0\textrm{ if }N_{2}^{*}>0,\lambda_{3}=0\textrm{ if }N_{2}^{*}<\frac{1}{b};\label{eq:kkt2}\\
\frac{d\mathcal{L}}{ds}=-Y_{1}^{-\sigma}+h(1-bN_{2}^{*})Y_{2}^{-\sigma}+\mu & =0;\label{eq:kkt3}\\
N_{1}^{*},N_{2}^{*},s^{*},\lambda_{1},\lambda_{2},\lambda_{3},\mu & \ge0;N_{2}^{*}\le\frac{1}{b}.
\end{align}
Note that the Inada condition ($\lim_{x\rightarrow0}u'(x)=\infty$)
for period 1 rules out $s^{*}=1$ and $N_{1}=1/b$, so we need not
impose these constraints explicitly. Also, so long as $N_{2}^{*}<1/b$,
the same condition rules out $s^{*}=0$. We consider four cases,
of which only three can occur.

@@

Conditions (\ref{eq:kkt1}) and (\ref{eq:kkt2}) give the bounds for
this case. When $\sigma<1$, (\ref{eq:kkt1}) puts a minimum on $h$
and (\ref{eq:kkt2}) puts a maximum on $h$. The situation is reversed
for $\sigma>1$.

@@

>>>>

Plugging (\ref{eq:Y1}) and (\ref{eq:Y2}) into the above, we can
derive the Karush-Kuhn-Tucker conditions for an optimum $(N_{1}^{*},N_{2}^{*},s^{*})$
as:
\begin{align}
\frac{d\mathcal{L}}{dN_{1}}=-bY_{1}^{-\sigma}+a+\lambda_{1} & =0\textrm{, with }\lambda_{1}=0\textrm{ if }N_{1}^{*}>0;\label{eq:kkt1}\\
\frac{d\mathcal{L}}{dN_{2}}=-bs^{*}hY_{2}^{-\sigma}+a+\lambda_{2} & -\lambda_{3}=0\textrm{, with }\lambda_{2}=0\textrm{ if }N_{2}^{*}>0,\lambda_{3}=0\textrm{ if }N_{2}^{*}<\frac{1}{b};\label{eq:kkt2}\\
\frac{d\mathcal{L}}{ds}=-Y_{1}^{-\sigma}+h(1-bN_{2}^{*})Y_{2}^{-\sigma}+\mu & =0;\label{eq:kkt3}\\
N_{1}^{*},N_{2}^{*},s^{*},\lambda_{1},\lambda_{2},\lambda_{3},\mu & \ge0;N_{2}^{*}\le\frac{1}{b}.
\end{align}
Note that the Inada condition (that marginal utility of income grows without 
bound as income approaches zero, $\lim_{x\rightarrow0}u'(x)=\infty$)
for period 1 rules out $s^{*}=1$ and $N_{1}=1/b$, so we need not
impose these constraints explicitly. Also, so long as $N_{2}^{*}<1/b$,
the same condition rules out $s^{*}=0$. We consider four cases,
of which only three can occur.

@@

This case holds for intermediate values on $h$. Equation (\ref{eq:dNdh2}) 
shows that for $\sigma < 1$, $N_2$ decreases in $h$; the requirement that
$N_2>0$ therefore puts a maximum on $h$. When $\sigma>1$, $N_2$ increases in
$h$ and this puts a minimum on $h$. The requirement $N_1 = 0$ provides the 
other bound. Equation (\ref{eq:kkt1}) 
requires $-bY_{1}^{-\sigma}+a \le 0$ since $\lambda_1$ must be non-negative. 
The LHS is increasing in $Y_1$, and hence decreasing in $s$ as $Y_1 = 1-s$ since 
$N_1=0$. Lastly, optimal choice of education $s^*$ increases in $h$ for 
$\sigma<1$, and decreases for $\sigma > 1$. Hence for $\sigma < 1$, (\ref{eq:kkt1})
puts a minimum on $h$, and for $\sigma > 1$ it puts a maximum on $h$.

@@

\subsection*{Effect of $a$}

\begin{lem}
\label{lemma-a}For $\sigma<1$, $d^2N^*/dadh > 0$, i.e. the effect of $a$
increases at higher levels of human capital.
\end{lem}
\begin{proof}
Differentiating (\ref{eq:dNdh1}) with respect to $a$ gives

\[
\frac{d^2N^*_t}{dadh} = -\frac{1}{b}\frac{1}{1-2\sigma}a^{-1}\left(\frac{a}{b}\right)^{1/(1-2\sigma)}\frac{1-\sigma}{2\sigma-1}h^{(1-\sigma)/(2\sigma-1)-1}
\]

for $t=1,2$ when $N^*_1,N^*_2>0$. For $\sigma > 0.5$ this is positive.

When $N^*_1 = 0,N^*_2>0$, differentiating (\ref{eq:dNdh2}) with respect to $a$ gives

\[
\frac{d^2N_{2}^{*}}{dadh} = \frac{1}{b}\frac{1}{\sigma}a^{-1}\left(\frac{a}{b}\right)^{-1/\sigma}\frac{1-\sigma}{\sigma}(s^{*}h)^{(1-2\sigma)/\sigma}(s^{*}+h\frac{ds^{*}}{dh})
\]

which is $-a^{-1}/\sigma$ times (\ref{eq:dNdh2}) and hence is positive.

\end{proof}
